**Evaluation of machine learning methods for 16S rRNA gene data**

### 2019-06-12 submission/manuscript.Rmd
figure axis labels bigger

We compared the median cross-validation AUROC and median testing AUROC for each model. This difference should be low to suggest the model is not overfitting despite the large number of features. The largest difference between the two was `r paste(round(abs(difference_model[difference_index[7],1]), 3))` in L1 SVM with linear kernel followed by SVM with radial basis function kernel and decision tree with a difference of `r paste(round(abs(difference_model[difference_index[6],1]), 3))` and `r paste(round(abs(difference_model[difference_index[5],1]), 3))`, respectively [Figure 2].

We reported the testing AUROC values over 100 random splits of the full dataset [Figure 1]. The testing AUROC values varied `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 2))` on average, depending on the data-split. For instance, the lowest AUROC value of the random forest model was `r paste(round(min((all %>% filter(model=="Random_Forest"))$test_aucs), 2))` whereas the highest was `r paste(round(max((all %>% filter(model=="Random_Forest"))$test_aucs), 2))`.

@@

The ML models we built using L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost decrease in interpretibility as they increase in complexity. We interpreted L1 and L2 SVM with linear kernel and L2 logistic regression using the feature weights of the trained models. We ranked the absolute weights of all the OTUs for each data-split. We calculated the median ranks of these features over the 100 data-splits. In the three linear models, OTUs that had the largest median ranks and drove the detection of SRNs belong to family *Lachnospiraceae*,  and *Ruminococcaceae* (OTU01239, OTU00659, OTU00742, OTU00012, OTU00015, OTU00768, OTU00822, OTU00609) and genera *Gamella*  (OTU00426) [Figure 3]. Some of the OTUs with the highest ranks were shared among the linear models. We explained the feature importances in non-linear models using permutation importance on the held-out test data where we randomly permuted non-correlated features individually and groups of correlated features together (see methods) to calculate the effect on testing AUROC. The top 5 OTUs with the largest negative impact on testing AUROC were overl in tree-based models [Figure 4]. Specifically, permuting *Peptostreptococcus* (OTU00367) abundances randomly, dropped the predictive performances the most for all tree-based methods. Decision tree, random forest and XGBoost models' predictive performance dropped from `r paste(round(dt_median, 2))` base testing AUROC median to `r paste(round(median((dt_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, from `r paste(round(rf_median, 2))` to `r paste(round(median((rf_imp %>% filter(names=="Otu00367"))$new_auc), 2))` and from `r paste(round(xgboost_median, 2))` to `r paste(round(median((xgboost_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, respectively [Figure 4]. The negative predictive impact of *Peptostreptococcus* in the decision tree model was followed by a *Lachnospiraceae* species (OTU00058) (`r paste(round(dt_median, 2))` base testing AUROC median to `r paste(round(median((dt_imp %>% filter(names=="Otu00058"))$new_auc), 2))`) [Figure 4B]. Other OTUs had none to minimal effect on the predictive performance.

@@

As the complexity of a ML model increase, its training times will increase (Table S1). When regularization and hyperparameter optimization (i.e. model selection) steps are added to the model pipeline, training times will increase further. The more hyperparameters a model has, the longer it will take to train. The computational burden that comes from model complexity, can be reduced by parallelizing different segments of the ML pipeline. In this study we have parallelized each datasplit which allowed 100 atasplits to be run at the same time for each model. We can also parallelize cross-validation hyerp-parameter search for each hyper-parameter setting.

>>>>

We compared the median cross-validation AUROC and median testing AUROC of each model to one another. This difference should be low to suggest the model is not overfitting despite the large number of features. The largest difference between the two was `r paste(round(abs(difference_model[difference_index[7],1]), 3))` in L1 SVM with linear kernel followed by SVM with radial basis function kernel and decision tree with a difference of `r paste(round(abs(difference_model[difference_index[6],1]), 3))` and `r paste(round(abs(difference_model[difference_index[5],1]), 3))`, respectively [Figure 2].

We reported the testing AUROC values over 100 random splits of the full dataset which showed that depending on the data-split, the testing AUROC value showed great variability [Figure 1]. The testing AUROC values within each model varied `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 2))` on average across the seven models. For instance, the lowest AUROC value of the random forest model was `r paste(round(min((all %>% filter(model=="Random_Forest"))$test_aucs), 2))` whereas the highest was `r paste(round(max((all %>% filter(model=="Random_Forest"))$test_aucs), 2))`.

@@

The ML models we built using L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost decrease in interpretibility as they increase in complexity. We interpreted L1 and L2 SVM with linear kernel and L2 logistic regression using the feature weights of the trained models. We ranked the absolute weights of all the OTUs for each data-split. We calculated the median ranks of these features over the 100 data-splits. In the three linear models, OTUs that had the largest median ranks and drove the detection of SRNs belong to family *Lachnospiraceae*,  and *Ruminococcaceae* (OTU01239, OTU00659, OTU00742, OTU00012, OTU00015, OTU00768, OTU00822, OTU00609) and genera *Gamella*  (OTU00426) [Figure 3]. Some of the OTUs with the highest ranks were shared among the linear models. We explained the feature importances in non-linear models using permutation importance on the held-out test data where we randomly permuted non-correlated features individually and groups of correlated features together (see methods) to calculate the effect on testing AUROC. The top 5 OTUs with the largest negative impact on testing AUROC were overlapped in tree-based models [Figure 4]. Specifically, permuting *Peptostreptococcus* (OTU00367) abundances randomly, dropped the predictive performances the most for all tree-based methods. Decision tree, random forest and XGBoost models' predictive performance dropped from `r paste(round(dt_median, 2))` base testing AUROC median to `r paste(round(median((dt_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, from `r paste(round(rf_median, 2))` to `r paste(round(median((rf_imp %>% filter(names=="Otu00367"))$new_auc), 2))` and from `r paste(round(xgboost_median, 2))` to `r paste(round(median((xgboost_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, respectively [Figure 4]. The negative predictive impact of *Peptostreptococcus* in the decision tree model was followed by a *Lachnospiraceae* species (OTU00058) (`r paste(round(dt_median, 2))` base testing AUROC median to `r paste(round(median((dt_imp %>% filter(names=="Otu00058"))$new_auc), 2))`) [Figure 4B].

@@

As the complexity of a ML model increase, its training times will increase as well (Table S1). When regularization and hyperparameter optimization (i.e. model selection) steps are added to the model pipeline, training times will increase further. The more hyperparameters a model has, the longer it will take to train. The computational burden that comes from model complexity can be reduced by parallelizing different segments of the ML pipeline. In this study we have parallelized each data-split which allowed 100 data-splits to be run at the same time for each model. We can further parallelize the cross-validation step for each hyper-parameter setting.
